---
title: "Create maps with R"
author: "Christophe Bontemps (UN SIAP)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes
  html_notebook: default
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: lumen
  word_document: default
---


```{r setup, include=FALSE}
# Remeber to disable png for Word export...
knitr::opts_chunk$set( message = FALSE, warning = FALSE, 
                       results =FALSE, echo = FALSE,
                       dev="png", 
                       dev.args=list(type="cairo"), dpi=96)

# My colors:
SIAP.color <- "#0385a8"

# Function used to recreate ggplot standard colors
ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

```


```{r libraries}
library(tidyverse)
library(viridis)
library(ggthemes)
library(ggplot2)
library(forcats)
library(knitr)
library(kableExtra)

```
# Useful sources {-}

* [**Fundamentals of Data Visualization**](https://clauswilke.com/dataviz/index.html) by Claus O. Wilke
* [**Geocomputation with R**](https://geocompr.robinlovelace.net/index.html) Robin Lovelace, Jakub Nowosad, Jannes Muenchow, a,d in particular the chapter [*Making maps with R*](https://geocompr.robinlovelace.net/adv-map.html) 
* Kieran Healy's [**Data Visualization book**](https://socviz.co/maps.html#maps) chapter on maps

## On ggplot 

[ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/maps.html) by Hadley Wickham


# Data sets {-}
We use the data from  [Natural Earth Data](https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-admin-0-countries/) for 
*all countries*, and in particular the sahpefiles proposed in the file *ne_10m_admin_0_countries.shp*



# Making maps with the *sf* library

*We use the tutorial provided by [Atlan](https://atlan.com/courses/introduction-to-gis-r/lesson2-manipulating-geospatial-data/)*


The *sf* (Simple Features) package is the successor to the *sp* package. It allows to work in the *tidy* framework using spatial dataframe. Therefore, one can use the pipe an otehr tidy tools to  manipulate an sf object like an ordinary dataframe, or at least one with a few special characteristics.


```{r Data}
library(sf)

#loading Shapefiles
library(sf)
my_sf <- st_read("Data/Maps/ne_10m_admin_0_countries/ne_10m_admin_0_countries.shp")

glimpse(my_sf)

```


```{r}
st_crs(my_sf)

```


> One can extract only the geometry from the Spatial data.frame to see the shapes of the map

```{r}
#oExtract only the geometry with sf:st_geometry

my_geometry <- st_geometry(my_sf)
plot(my_geometry, main = "Map Geometry")


```

There are many information already in the shapefile, we can plot all of them here...

```{r}
plot(my_sf)
```

... or just one... 

```{r}
plot(my_sf["POP_EST"], 
     key.pos = 1,
     key.width = lcm(1.3), 
     key.length = 0.5,
     main = "Population (est. 2017)")
```

... or part of it..

```{r}
my_asia <- my_sf %>%
  filter(CONTINENT == "Asia" | CONTINENT == "Oceania" )

plot(my_asia["POP_EST"], 
     key.pos = 1,
     key.width = lcm(1.3), 
     key.length = 0.5,
     main = "Population (est. 2017)")

```


## Merging Spatial data sets

Here we already have a spatial data object that has its own projection system ( `r st_crs(my_sf)[1]`). We will merge this  spatial data frame  with a data set where the SDG indicator of interest (5.5.1) is recorded at the at the country level. The merging variable will be the name of the country, so there is no need here to worry about the projection correspondance and the adjustment of the spatial units reference since only one file is using a projection system.   

```{r}

my_sdg <- read.csv("Data/Maps/SG_GEN_PARL_ap.csv", sep = ",")
sdg_label <- "Population of seats held by women in national parliaments"

# Keeping only esential values in the ShapeFile dataset
# my_asia_reduced <- my_asia  %>%
#   select("ADMIN", "NAME", "NAME_LONG")
  

my_map <- merge(my_sf, my_sdg, by.x= "NAME_LONG", by.y = "GeoAreaName")  %>%
  st_crop(c(xmin= -50, xmax = 500, ymin = -50,  ymax = 70))


```


```{r}
plot(my_map["X2019"], 
     key.pos = 1,
     key.width = lcm(1.3), 
     key.length = 0.5, 
     main = sdg_label, 
     sub = "(SDG indic. 5.5.1)")
```

## Adjusting the breaks
Here we clearly see that the colors applied to the map do not help understanding the various *levels$* of percentages for each country. 
We can change the numbers of categories, the colors and length of each color category used for doing this map.  
```{r}
# breaks using quintiles 

my_nb_breaks <- 5

plot(my_map["X2019"], 
     nbreaks = my_nb_breaks,
     breaks = "jenks",
     pal = heat.colors(my_nb_breaks),
     key.pos = 1,
     key.width = lcm(1.4), 
     key.length = 0.5, 
     main = sdg_label
     )

```



# With ggplot 

ggplot can easily represent an *sf* object and represent it as it would for any data frame using the **geom_sf**

```{r}
library(ggplot2)
ggplot(my_data) + 
  geom_sf(aes(fill = X2019)) +
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")
```

```{r}
 ggplot(my_data) + 
  geom_sf(aes(fill = X2019)) +
  ggtitle(sdg_label) +
  theme_minimal() +
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme(legend.position = "bottom", 
        legend.title = element_text(color = "#235a91", size = 10))
```

## Changing the map projection 

```{r}
ggplot(my_data) + 
  geom_sf(aes(fill = X2019)) +
  ggtitle(sdg_label) +
  theme_minimal() +
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme(legend.position = "bottom", 
        legend.title = element_text(color = "#235a91", size = 10)) +
  coord_sf(crs = st_crs(3112))
  
```




```{r}
## regions Colors
#my_cols <- c("#00AFBB", "#E7B800", "#FC4E07") 
#my_cols <-c("#6b03fc", "#4287f5", "#69b3a2", "#03fca1", "#dffc03", "#E8E8E8") )
my_cols <- c(ggplotColours(n = 5), "#E8E8E8") 
my_cols <- c("#F8766D","#A3A500" , "#69b3a2","#00B0F6", "#E76BF3", "#E8E8E8")

library(RColorBrewer)
# Tree colors
my_highlow <- c('#fc8d59','#ffffbf','#91bfdb')
```



# With *tmap*

```{r}
knitr::knit_exit()
```





```{r DataLabelCleaning}
# Cleaning variables names
library(janitor)
LongLabels <- names(WV) # keeping for future usage

names(WV) <-Shortlabels
WV <-  janitor::clean_names(WV)

names(WV)[8] <-  "PPP Income per cap"
names(WV)[10] <- " GDP growth per cap "
names(WV)[11] <- " Share of Income (<20%)"
names(WV)[12] <- " Child malnutrition (%)"
names(WV)[15] <- " HIV (per 1000)"
names(WV)[16] <- " Tuberculosis (per 100k)"
names(WV)[17] <- " Road Mortality (per 100k)"
names(WV)[19] <- " Male employment (%)"
names(WV)[20] <- " Female employment (%)"
names(WV)[22] <- " Pop. Access Water (%)"
names(WV)[23] <- " Pop. Access Sanitary (%)"
names(WV)[27] <- " Slums Pop.(%)"
names(WV)[29] <- " Savings (% of GNI)"
names(WV)[30] <- " CO2 per cap."
names(WV)[31] <- " Marine areas (%)"
names(WV)[48] <- " Developped"

# removing  variables (country classification)
WV <- WV %>%
  select(!c(34:35, 40:43, 45:47, 49))

var.select <- c(4, 8, 10:33)
var.shortlist <-c(8, 10:16, 18, 20:22, 29:30)

```

### Glimpse of the data{-}

```{r TableFull, echo = FALSE, results=TRUE}
t <-  kbl(WV[1:10,c(1,var.shortlist)])
kable_styling(t,full_width = F)

# For interactive view of recording 
# View(SDGPov[c("Indicator","Age", "Poverty" )])
```


# Missing values 

```{r missing}
library(naniar)
# See https://cran.r-project.org/web/packages/naniar/vignettes/naniar-visualisation.html
# miss_var_summary(WV, order = TRUE)

gg_miss_var(subset(WV, is.na(region_name)== FALSE),
            show_pct = TRUE) + 
  labs(y = "Missing observations (%)") +
  theme(
    plot.title = element_text(size=10), 
    axis.text.y =  element_text(size=4)
  )

```

## By region{-}
```{r missing1}
library(naniar)
# See https://cran.r-project.org/web/packages/naniar/vignettes/naniar-visualisation.html
# miss_var_summary(WV, order = TRUE)

gg_miss_var(subset(WV, is.na(region_name)== FALSE),
            show_pct = TRUE, 
            facet = region_name ) + 
  labs(y = "Missing observations (%)") +
  theme(
    plot.title = element_text(size=10), 
    axis.text.y =  element_text(size=4)
  )

```


## Heatmap


```{r missingHeat}
Mytemp <- subset(WV[,], is.na(region_name)== FALSE)

# From https://github.com/njtierney/naniar/issues/183 
ggmissfct <- function(order_missing = FALSE, custom_order = NULL) {
    if (order_missing == TRUE) {
        desc_missing <- Mytemp %>%
            dplyr::select(-region_name) %>% 
            miss_var_summary() %>% 
            dplyr::pull(variable)
    } else {
        desc_missing <- NULL
    }
    
    Mytemp %>%
        dplyr::group_by(region_name) %>%
        miss_var_summary() %>%
        dplyr::mutate(variable = factor(variable,
                                        levels = (sort(unique(variable),
                                                       decreasing = TRUE))) %>% 
                          forcats::fct_relevel(rev(desc_missing), after = Inf) %>% 
                          forcats::fct_relevel(rev(custom_order), after = Inf)) %>%
        ggplot(aes(region_name,
                   variable,
                   fill = pct_miss)) +
        geom_tile() +
        viridis::scale_fill_viridis(name = "% Miss") +
        theme(
    plot.title = element_text(size=10),
    legend.position = "bottom",
    axis.text.y =  element_text(size=5)
  )
    
}

# By default
ggmissfct(order_missing = FALSE)
```

```{r missingHeat2}
#Ordered by missing percentage
ggmissfct(order_missing = TRUE)
```

## Imputing missing values (for a better visual demonstration){-}


```{r missing3}
library(Amelia)
# Computing values to input (long computation)
#res.amelia <- amelia(WV[,2:33], m = 2)  
#save(res.amelia, file ="Data/res.amelia")

load(file = "Data/res.amelia")

# Binding the imputed variables 
WV.comp <- cbind(WV[,1], res.amelia$imputations$imp1, WV[,34:ncol(WV)])
names(WV.comp)[1] <- "country"

missmap(WV.comp, 
        rank.order = TRUE)

```


# Continuous and discrete

## Histogram 

```{r HistoSpineGrey}
WV %>%
 filter(!is.na(` CO2 per cap.`)) %>%
 filter(!is.na(region_name)) %>%
 filter(` Developped` %in% 
    c("Developing", "Developed") | is.na(` Developped`)) %>%
 ggplot() +
 aes(x = ` CO2 per cap.`  ) +
 geom_histogram(bins = 61L, fill = "#bdbdbd") +
 theme_minimal() +
 theme(legend.position = "bottom")

```

# Rather than stacking histograms, it's easier to compare frequency


```{r HistoSpine}
WV %>%
 filter(!is.na(` CO2 per cap.`)) %>%
 filter(!is.na(region_name)) %>%
 filter(` Developped` %in% 
    c("Developing", "Developed") | is.na(` Developped`)) %>%
 ggplot() +
 aes(x = ` CO2 per cap.`, fill = region_name) +
 geom_histogram(bins = 61L) +
 scale_fill_hue() +
 theme_minimal() +
 theme(legend.position = "bottom")

```

# Categories

```{r CatData}
# We create new categorical variables
# Computing quantiles  of CO2
Q_CO2=quantile(WV$` CO2 per cap.`,c(0,1/3,2/3,1), na.rm = TRUE)
Q_CO2[1]=Q_CO2[1]-.00005

# Computing quantiles  of population
Q_pop=quantile(WV$population,c(0,1/4,1/2, 3/4, 1), na.rm = TRUE)
Q_pop[1]=Q_pop[1]-.00005

# Computing quantiles  of population
Q_pop_dens=quantile(WV$population_density,c(0,1/4,1/2, 3/4, 1), na.rm = TRUE)
Q_pop_dens[1]=Q_pop[1]-.00005

WV <- WV %>%
  mutate(
    CO2_emissions_cat = cut(` CO2 per cap.`,  breaks=Q_CO2, 
                            labels=c("low","medium","high")),
    Country_pop_cat = cut(population,  breaks=Q_pop, 
                      labels=c("lower", "low","high", "highest")),
    
    Country_density_cat = cut(population_density,  breaks=Q_pop_dens, 
                      labels=c("lower", "low","high", "highest")),

  )
#names(WV)[ncol(WV)] <- "CO2 emission per cap. (cat.) "

```

## Mosaic plot

```{r mosaic}


mosaicplot(table(WV$region_name, WV$CO2_emissions_cat),
  #color = TRUE,
  #shade = FALSE,
   col = rev(my_highlow), 
  main = " Mosaic plot of CO2 emisions per region", 
  xlab = "Regions", # label for x-axis
  ylab = paste("CO2 emissions \n (High Medium, Low)") # label for y-axis
)
```

## Stacked bars

```{r StackedBar}
WV %>%
 filter(
   !is.na(` CO2 per cap.`),
   !is.na(region_name),
   ` Developped` %in%   c("Developing", "Developed") | is.na(` Developped`)) %>%
 ggplot() +
 aes(x = CO2_emissions_cat ,fill = region_name, weight = population) +
 geom_bar() +
 scale_fill_hue() +
labs( title= "Population by region according to CO2 emissions classification \n(Stacked bars)" ) +
 ylab("Population (millions)")+
 xlab(" CO2 emissions category")+
 theme_tufte() +
 theme(legend.position = "bottom")
```



```{r StackedBarDodged}
WV %>%
 filter(
   !is.na(` CO2 per cap.`),
   !is.na(region_name),
   ` Developped` %in%   c("Developing", "Developed") | is.na(` Developped`)) %>%
 ggplot() +
 aes(x = CO2_emissions_cat ,fill = region_name, weight = population) +
 geom_bar(position = "dodge") +
 labs( title= "Population by region according to CO2 emissions classification \n(dodged bars)" ) +
 ylab("Population (millions)")+
 xlab("CO2 emissions category")+
 scale_fill_hue() +
 theme_tufte() +
 theme(legend.position = "bottom")
```


```{r StackedBarSmall}
WV %>%
 filter(
   !is.na(` CO2 per cap.`),
   !is.na(region_name),
   ` Developped` %in%   c("Developing", "Developed") | is.na(` Developped`)) %>%
 ggplot() +
 aes(x = CO2_emissions_cat ,fill = CO2_emissions_cat, weight = population) +
 geom_bar(position = "dodge") +
 facet_wrap(~region_name) +
 labs( title= "Population by region according to CO2 emissions classification \n(small multiple bars)" ) +
ylab("Population (millions)")+
 xlab("CO2 emissions category")+
 scale_fill_brewer(palette = "Oranges") +
 theme_tufte() +
 theme(legend.position = "bottom")
```

# Dual version, looking at CO2 emissions by population density categories

```{r StackedCO2}
WV %>%
 filter(
   !is.na(Country_density_cat),
   !is.na(region_name),
   ` Developped` %in%   c("Developing", "Developed") | is.na(` Developped`)) %>%
 ggplot() +
 aes(x = Country_density_cat, fill = region_name, weight = ` CO2 per cap.`) +
 geom_bar() +
 ggtitle("CO2 emissions per population density classification of countries (stacked bars)" ) +
 ylab("CO2 emissions per cap.")+
 xlab("Countries density levels")+
 scale_fill_hue() +
  theme_tufte() +
 theme(legend.position = "bottom")
```



```{r StackedCO2Dogde}
WV %>%
 filter(
   !is.na(Country_density_cat),
   !is.na(region_name),
   ` Developped` %in%   c("Developing", "Developed") | is.na(` Developped`)) %>%
 ggplot() +
 aes(x = Country_density_cat, fill = region_name, weight = ` CO2 per cap.`) +
 geom_bar(position = "dodge") +
 ggtitle("CO2 emissions per population density classification of countries (dodged bars)" )+
 ylab("CO2 emissions per cap.")+
 xlab("Countries density levels")+
 scale_fill_hue() +
  theme_tufte() +
 theme(legend.position = "bottom")
```
# Scatterplot 

```{r ScatterSize}

S <-WV %>%
 filter(` Developped` %in% c("Developing", "Developed") | is.na(` Developped`),
        population_density < 5000, 
        !is.na(region_name)) %>%
 ggplot() +
 aes(x = population_density, y = ` CO2 per cap.`, colour = region_name, size = population) +
 geom_point() +
 ggtitle(" Relationship between CO2 emissions and country density") +
 ylab("CO2 emissions per cap.")+
 xlab("Countries density (people per sq. km)")+
 theme_tufte() +
 theme(legend.position = "bottom", legend.box="vertical")

S

```

## Interactive version 

```{r ScatterPlotly, eval= FALSE}
library(plotly)
ggplotly(S)
style(S, hoverinfo = "none", traces = 1:2)
```
# Small multiple


```{r Scattermultiple}
WV %>%
 filter(` Developped` %in% c("Developing", "Developed") | is.na(` Developped`),
        population_density < 5000, 
        !is.na(region_name)) %>%
 ggplot() +
 aes(x = population_density, y = ` CO2 per cap.`, colour = region_name) +
 geom_point() +
 facet_wrap(~region_name)+
 ggtitle(" Relationship between CO2 emissions and country density per region") +
 ylab("CO2 emissions per cap.")+
 xlab("Countries density (people per sq. km)")+
 theme_tufte() +
 theme(legend.position = "bottom")
```



# Scatterplot Matrix simple

```{r Scattermatrixsmall}

pairs(subset(WV.comp[,c(10, 12,15, 30)], !is.na(WV.comp$region_name)),
             , pch = 19, 
      col = my_cols[WV$region_name],
      #col = "grey",
      upper.panel = NULL)
```

# Scatterplot Matrix full


```{r Scattermatrix}

pairs(WV.comp[,var.shortlist], pch = 19, 
       col = my_cols[WV$region_name],
      # col = "grey",     
      upper.panel = NULL,
      cex.labels = 0.3)
```




# Combined Scatterplot/correlation with GGally::ggpairs

```{r ggpairs1}
library(GGally)
WVshort <- WV.comp[,c(10,12,15,30,37)] %>%
  filter(
  !is.na(WV.comp$region_name)
  )
 
ggpairs(WVshort,
        columns = 1:4,
        columnLabels = gsub('.', ' ', colnames(WVshort[,1:4]), fixed = T), 
        labeller = label_wrap_gen(10), 
        upper = "blank",
        #upper = list(continuous = wrap("cor", size = 9)),
        aes(colour = WVshort$region_name, alpha= 0.1),
        title="Correlogram for 4 variables") +
 theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=9, angle = 60, hjust = 1), 
    #axis.text.x =  element_blank(),
    legend.position = "bottom"
  )


```


```{r ggpairs2}
WVshort <- WV.comp[, c(var.shortlist[1:10], 37)] 
 
ggpairs(WVshort,
        columns = 1:10,
        columnLabels = gsub('.', ' ', colnames(WVshort[,1:10]), fixed = T), 
        labeller = label_wrap_gen(10), 
        upper = "blank",
        lower = list(continuous = wrap("points", alpha = 0.3, size=0.2)),
        #upper = list(continuous = "cor"),
        aes(colour = WV$region_name, alpha= 0.1 ), 
        title="Correlogram for 10 variables") +
 theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=9, angle = 60, hjust = 1), 
    #axis.text.x =  element_blank(),
    legend.position = "bottom"
  )


```



```{r ggpairs3,  eval = FALSE}
WVshort <- WV.comp[, c(2:32, 37)] 
 
gg <-ggpairs(WVshort,
        columns = 1:31,
        columnLabels = gsub('.', ' ', colnames(WVshort[,1:31]), fixed = T, ), 
        labeller = label_wrap_gen(10), 
        lower = list(continuous = wrap("points", alpha = 0.3, size=0.05)),
        upper = "blank",
        #upper = list(continuous = wrap("cor", size = 9)),
        aes(colour = WVshort$region_name, alpha= 0.1),
        title="Correlogram for 30+ variables")

gg + theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=4, angle = 60, hjust = 1), 
    axis.text.y =  element_text(size=4), 
    #axis.text.x =  element_blank(),
    legend.position = "bottom"
  )


```







# Correlation Matrix

```{r CorrelationMatrix1}
library(ggcorrplot)

# We compute the correlation matrix of the covariates
corr_coef<-cor(WV.comp[,c(30, 10, 12,15,18, 22)],use = "p")

#And then plot it with nice options 
ggcorrplot(corr_coef, 
           type = "lower",         # lower triangle of the matrix only
           #hc.order = TRUE,        # variable sorted from highest to lowest
           outline.col = "white",  #Color options
           lab = TRUE)+
  theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=9, angle = 60, hjust = 1), 
    #axis.text.x =  element_blank(),
    #legend.position = "bottom"
  )

```


```{r CorrelationMatrix2}
library(ggcorrplot)

# We compute the correlation matrix of the covariates
corr_coef<-cor(WV.comp[, 2:33],use = "p")

#And then plot it with nice options 
ggcorrplot(corr_coef, 
           type = "lower",         # lower triangle of the matrix only
           # hc.order = TRUE,        # variable sorted from highest to lowest
           outline.col = "white",  #Color options
           lab = FALSE)+
  theme_tufte() +
   theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=6, angle = 60, hjust = 1),
    legend.title = element_text(size=8),
    legend.key.size = unit(0.5,"line")
    #axis.text.x =  element_blank(),
    #legend.position = "bottom"
  )
```



```{r CorrelationMatrix3}
library(ggcorrplot)

# We compute the correlation matrix of the covariates
corr_coef<-cor(WV.comp[, 2:33],use = "p")

#And then plot it with nice options 
ggcorrplot(corr_coef, 
           type = "lower",         # lower triangle of the matrix only
           hc.order = TRUE,        # variable sorted from highest to lowest
           outline.col = "white",  #Color options
           lab = FALSE)+
  theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=6, angle = 60, hjust = 1),
    legend.title = element_text(size=8),
    legend.key.size = unit(0.5,"line")
    #axis.text.x =  element_blank(),
    #legend.position = "bottom"
  )
```

# Parallel Coordinates chart

```{r ParallelMass}
library(MASS)

#parcoord(WV[,c(4,6, 8, 10)], col = "grey")
parcoord(WV.comp[,var.shortlist], col = "grey", 
         var.label = TRUE)


```


```{r ParallelGblack }
library(GGally)
library(viridis)
library(hrbrthemes)
ggparcoord(WV.comp[,c(var.shortlist, 35)],
    columns = c(1: length(var.shortlist)),
    #groupColumn = "region_name",
    showPoints = FALSE, 
    scale="uniminmax",    # "std" "center",
    #missing = "median",  # Not needed since WV.com
    title = "Parallel Coordinate Plot for Global Economy variables",
    alphaLines = 0.9
    ) + 
 # scale_color_manual(values=c("#69b3a2", "#E8E8E8", "#E8E8E8", "#E8E8E8", "#E8E8E8")) +
 # scale_color_viridis(discrete=TRUE) +
 theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=9, angle = 60, hjust = 1), 
    #axis.text.x =  element_blank(),
    legend.position = "bottom"
  )


```




```{r ParallelGrey }
ggparcoord(WV.comp[,c(var.shortlist, 35)],
    columns = c(1: length(var.shortlist)),
    #groupColumn = "region_name",
    showPoints = FALSE, 
    scale="uniminmax",    # "std" "center",
    #missing = "median",  # Not needed since WV.com
    title = "Parallel Coordinate Plot for Global Economy variables",
    alphaLines = 0.3
    ) + 
 # scale_color_manual(values=c("#69b3a2", "#E8E8E8", "#E8E8E8", "#E8E8E8", "#E8E8E8")) +
 # scale_color_viridis(discrete=TRUE) +
 theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=9, angle = 60, hjust = 1), 
    #axis.text.x =  element_blank(),
    legend.position = "bottom"
  )


```


```{r ParallelG1 }
p<-ggparcoord(WV.comp[,c(1, var.shortlist, 35)],
    columns = c(2: length(var.shortlist)),
    groupColumn = "region_name",
    showPoints = FALSE, 
    scale="uniminmax",    # "std" "center",
    #missing = "median", 
    title = "Parallel Coordinate Plot for Global Economy variables",
    alphaLines = 0.3
    ) + 
  scale_color_manual(values=my_cols) +
  geom_line(aes(text= paste( country, "-", region_name)))+
 # scale_color_viridis(discrete=TRUE) +
 theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=8, angle = 60, hjust = 1), 
    #axis.text.x =  element_blank(),
    legend.position = "bottom"
  )
p
```

# Interactive version

```{r ParallelPlotly}
library(plotly)
ggplotly(p, 
         tooltip = c("country,region_name"),
         alpha = 0.6)

```


```{r ParallelSpecific}
ggparcoord(WV.comp,
    columns = var.shortlist,
    groupColumn = "region_name",
    showPoints = FALSE, 
    scale="uniminmax",
    missing = "median",
    title = paste("Parallel Coordinate Plot for Global Economy variables", "\n",
                  "Focusing on a specific region"),
    alphaLines = 0.9
    ) + 
 scale_color_manual(values=c( "#F8766D", "#E8E8E8", "#E8E8E8", "#E8E8E8", "#E8E8E8")) +
 # scale_color_viridis(discrete=TRUE) +
 theme_tufte() +
  theme(
    plot.title = element_text(size=10), 
    axis.text.x =  element_text(size=6, angle = 60, hjust = 1), 
    #axis.text.x =  element_blank(),
    legend.position = "bottom"
  )

```

# Another interactive version (axes reordering)

```{r parcoords }
library(parcoords)
# see https://cran.r-project.org/web/packages/parcoords/vignettes/introduction-to-parcoords-.html

# Here we can play with axis order

parcoords(WV.comp[, c(35, var.shortlist)],
  rownames = FALSE,
  #brushMode = '1D-axes',
  brushPredicate = "or", # "and" "or"
  alphaOnBrushed = 0.2,
  color = SIAP.color, alpha = 0.5,
  bundleDimension = " GDP growth per cap ",  # One variable
  bundlingStrength = 1, 
  smoothness = 0,  # lines = 0, curves = 1
  reorderable = TRUE)
```


```{r parcoordsBrush }
library(parcoords)
# see https://cran.r-project.org/web/packages/parcoords/vignettes/introduction-to-parcoords-.html

# Here we can play with brushing  and selecting (mainly)
parcoords(WV.comp[, c(35, var.shortlist)],
  rownames = FALSE,
  brushMode = '1D-axes',
  brushPredicate = "or", # "and" "or"
  alphaOnBrushed = 0.2,
  color = SIAP.color, alpha = 0.5,
  bundleDimension = " GDP growth per cap ",  # One variable
  bundlingStrength = 1, 
  smoothness = 0,  # lines = 0, curves = 1
  reorderable = TRUE)
```

## Paralel coordinates with soft lines

```{r parcoordssmooth }
library(parcoords)
# see https://cran.r-project.org/web/packages/parcoords/vignettes/introduction-to-parcoords-.html


parcoords(WV.comp[, c(35, var.shortlist)],
  rownames = FALSE,
  brushMode = '1D-axes',
  brushPredicate = "or", # "and" "or"
  alphaOnBrushed = 0.2,
  color = SIAP.color, alpha = 0.2,
  bundleDimension = " GDP growth per cap ",  # One variable
  bundlingStrength = 1, 
  smoothness = 0.2,  # lines = 0, curves = 1
  reorderable = TRUE)
```



```{r}
knitr::knit_exit()
```

