---
title: "Interactive Graphics with R"
author: "Christophe Bontemps (UN SIAP)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes
  html_notebook: default
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: lumen
  word_document: default
---


```{r setup, include=FALSE}
# Remeber to disable png for Word export...
knitr::opts_chunk$set( message = FALSE, warning = FALSE, 
                       results =FALSE, echo = TRUE,
                       dev="png", 
                       dev.args=list(type="cairo"), dpi=96)

# My colors:
SIAP.color <- "#0385a8"

# Function used to recreate ggplot standard colors
ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

```


```{r libraries}
library(knitr)
library(tidyverse)
library(forcats)
library(ggplot2)
library(ggthemes)
library(kableExtra)
library(viridis)

```
# Useful sources {-}


### On ggplot{-}

[ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/maps.html) by Hadley Wickham, and the [*sf* vignette](https://cran.r-project.org/web/packages/sf/vignettes/sf5.html)


## Data sets {-}

* We use data from [Gapminder website](https://www.gapminder.org/data/). For each of 142
countries, we collected values for CO2 emmissions (tonnes per person)  and Income per person (GDP per capita, adjusted in 2011 USD)
*The data sets can also be downloaded from the course LMS*


```{r}
Data_Income <- read.csv(file ="Data/M6-GapminderIncome.csv") %>%
   pivot_longer(!country, names_to = "year", values_to = "Income") %>%
  mutate(
    year = str_sub(year,2)
  )%>%
  filter(
    year >= 1990
  )

Data_CO2<-  read.csv(file ="Data/M6-GapminderCO2.csv")  %>%
  pivot_longer(!country, names_to = "year", values_to = "CO2")%>%
  mutate(
    year = str_sub(year,2)
  )%>%
  filter(
    year >= 1990
  )

Data_Pop<-  read.csv(file ="Data/M6-GapminderPopulation.csv")%>%
  pivot_longer(!country, names_to = "year", values_to = "Population")%>%
  mutate(
    year = str_sub(year,2)
  )%>%
  filter(
    year >= 1990
  )

GapminderData = merge(Data_Income, Data_CO2, by = c("country", "year"))
GapminderData = merge(GapminderData, Data_Pop, by = c("country", "year"))


CountryClassification <- read.csv("Data/UNSDCountriesMethodology.csv") %>%
  select(
    Country.or.Area, Region.Name, Sub.region.Name
  ) %>%
  rename(
    Region = Region.Name
  ) 
MyData = merge(GapminderData, CountryClassification, by.x = "country", by.y= "Country.or.Area")


```



```{r}

library(ggplot2)

ggplot(MyData) +
 aes(x = Income) +
 geom_histogram(bins = 59L, fill = SIAP.color, color = "white") +
 scale_y_continuous( limits = c(0, 70))+
 theme_minimal() +
 theme(legend.position = "top") 
#+  facet_wrap(vars(Region), nrow = 2L)

```




```{r}
p <- MyData %>%
 ggplot( aes(Income, CO2, size = Population, color=Region)) +
 geom_point( alpha = 0.2) +
 #geom_smooth(span = 0.75) +
 scale_color_brewer(palette = "Set3") +
 labs(x = "Income (GDP per capita)", 
      y = "CO2 emissions (tonnes per capita) ",
      title = "CO2 emission and Income over time",
      caption = "Source: Gapminder", 
      color = "Region",
      size = "Population ") +
 theme_minimal() +
 theme(legend.position = "bottom")
  
p
```


# Interactivty with plotly




```{r}
library(plotly)

ggplotly(p)

```



# Shiny

Shiny will need the data set, so let's save it in the directory of the Shiny app called *"MyFirstShiny"*
It is quite mandatory to have the [Shiny Cheat sheet](https://shiny.rstudio.com/images/shiny-cheatsheet.pdf) at hand


```{r}
write.csv(MyData, file ="Data/M6-IncomeCO2-Data.csv", sep = ";")
```


```{r}
knitr::knit_exit()
```




First, let us extract the R  code from this file  
```{r}

purl("M6-InteractivedataVisualization.Rmd")
```























## The Spatial data Frame (shape file)

The original file contains many variables with administrative information,some statistics (population, GDP, income, ... ) that are not useful for our work. Moreover, we can reduce the size of the map to include only the countries that we focus on. 

```{r, results = "asis"}
# We can only keep the variables and region we need...
SFvariables <- c("NAME_LONG", "ABBREV", "ADM0_A3", "REGION_UN",  "geometry" )

my_sf_asia <- my_sf %>%
  filter(CONTINENT == "Asia" | CONTINENT =="Oceania"
         ) %>%
  select(SFvariables)


t <-  kbl(my_sf_asia[1:10,])
kable_styling(t,full_width = F, )

```


```{r}

# Merging the geo spatial data wit SDG data at the country level

my_geodata <- merge(my_sf_asia, my_sdg, by.x= "NAME_LONG", by.y = "GeoAreaName") %>%
  st_crop(c(xmin= 00, xmax = 500, ymin = -50,  ymax = 70))

```

After merging at the country level we have now **in  the same spatial data frame** the country shape files and the SDG indicator we want to represent, plus some additional variables such as names, regions etc...


# Choropleth Maps 

## with *sf*

```{r}
plot(my_geodata["X2019"], 
     key.pos = 1,
     key.width = lcm(1.3), 
     key.length = 0.5, 
     main = sdg_label, 
     sub = "(SDG indic. 5.5.1)")
```

### Adjusting the breaks{-}

Here we clearly see that the colors applied to the map do not help understanding the various *levels* of percentages for each country. 

We can change the numbers of categories, the colors and length of each color category used for doing this map.  
There are various methods one can use to select the right color classes (manually, linear, quantiles, ...). Among the various methods, we have the *Jenks*the method seeks to reduce the variance within classes and maximize the variance between classes)

```{r}
# Breaks using different methods equal, quantile jenks

# Automatically
my_breaks <- "equal"
my_nb_breaks <- 10

# Manually
# my_breaks <- c(0,2, 4, 8, 16, 32, 40 )
# my_nb_breaks <- length(my_breaks) -1

plot(my_geodata["X2019"], 
     nbreaks = my_nb_breaks,
     breaks = my_breaks,
     # pal = heat.colors(my_nb_breaks),
     pal = sf.colors(my_nb_breaks, categorical = FALSE),
     key.pos = 1,
     key.width = lcm(1.4), 
     key.length = 0.5, 
     main = sdg_label
     )

```


## With ggplot 

ggplot can easily represent an *sf* object and represent it as it would for any data frame using the **geom_sf**. By default, since the variable is continuous, *ggplot* uses a **continuous color palette**.


```{r}
library(ggplot2)
ggplot(my_geodata) + 
  geom_sf(aes(fill = X2019)) +
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")
```
### A few improvements... {-}

```{r}
 ggplot(my_geodata) + 
  geom_sf(aes(fill = X2019)) +
  ggtitle(sdg_label) +
  theme_minimal() +
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme(legend.position = "bottom", 
        legend.title = element_text(color = "#235a91", size = 10))
```

### Changing the map projection {-} 

here we switch to Australia Lambert projection. 

```{r}
ggplot(my_geodata) + 
  geom_sf(aes(fill = X2019)) +
  ggtitle(sdg_label) +
  theme_minimal() +
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme(legend.position = "bottom", 
        legend.title = element_text(color = "#235a91", size = 10)) +
  coord_sf(crs = st_crs(3112))
  
```

### Changing the break points and number of categories...




#### Break points can be based on quantiles...{-}

```{r}
# Defining the nb of classes and breakpoints
nb_classes <- 8
```

Here with `r nb_classes` classes

```{r}
my_breaks <- quantile(my_geodata$X2019,
                      probs = seq(0, 1, length.out = nb_classes + 1),
                      na.rm = TRUE)
# Customize the labels
my_breaks_labels <-  c()
for(idx in 1:length(my_breaks)){
my_breaks_labels <- c(my_breaks_labels, paste0(round(my_breaks[idx], 0), 
                             " – ", 
                             round(my_breaks[idx + 1], 0)))
}
# I need to remove the last label 
# because that would be something like "66.62 - NA"
my_breaks_labels <- my_breaks_labels[1:length(my_breaks_labels)-1]

## We have to trasform the percentatges into a categorical variable. 
my_geodata %>%
  mutate(
    my_class = cut(X2019, 
                   breaks = my_breaks,
                   labels = my_breaks_labels, 
                   include.lowest = T))   %>%
ggplot() + 
  geom_sf(aes(fill = my_class)) +
  # We may use a different color palette
  #scale_fill_viridis(discrete = T)+
  ggtitle(sdg_label) + 
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme_minimal() +
  theme(legend.position = "bottom", 
        legend.title = element_text(color = "#235a91", size = 10))
  
```

#### Break points can be based on anything else {-}

```{r}
# Defining the nb of classes and breakpoints
nb_classes <- 8
my_breaks <- quantile(my_geodata$X2019,
                      probs = seq(0, 1, length.out = nb_classes + 1),
                      na.rm = TRUE)

my_breaks <-  seq(0, max(my_geodata$X2019, na.rm=T),
                  length.out = nb_classes + 1)
                    

# Customize the labels (rounding basically)
my_breaks_labels <-  c()
for(idx in 1:length(my_breaks)){
my_breaks_labels <- c(my_breaks_labels, paste0(round(my_breaks[idx], 0), 
                             " – ", 
                             round(my_breaks[idx + 1], 0)))
}
my_breaks_labels <- my_breaks_labels[1:length(my_breaks_labels)-1]

## We have to trasform the percentatges into a categorical variable. 
my_geodata %>%
  mutate(
    my_class = cut(X2019, 
                   breaks = my_breaks,
                   labels = my_breaks_labels, 
                   include.lowest = T))   %>%
ggplot() + 
  geom_sf(aes(fill = my_class)) +
  # We may use a different color palette
  #scale_fill_viridis(discrete = T)+
  ggtitle(sdg_label) + 
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme_minimal() +
  theme(legend.position = "bottom", 
        legend.title = element_text(color = "#235a91", size = 10))
  
```

> Cartography packages:  *tmap* and *cartography* {-}


## With the amazing library *tmap*

```{r}
library(tmap)

tmap_mode("view")
tm_shape(my_geodata) +
  tm_polygons("X2019") +
  tm_facets(by = "REGION_UN", as.layers = TRUE, free.coords = FALSE, drop.units = TRUE) 


```

```{r}
tmap_mode("plot")
tm_shape(my_geodata) +
  tm_polygons("X2019") +
  tm_facets(by = "REGION_UN")
```


```{r}
tmap_mode("view")
tm_shape(my_geodata) +
  tm_polygons("X2019") +
  tm_facets(by = "REGION_UN", as.layers = TRUE, free.coords = FALSE, drop.units = TRUE) 
```




### with the *Cartography* package 

```{r}
library(cartography)

# First: define the shape of the map
plot(st_geometry(my_geodata),
     col= "#f0f0f0",
     border = "black", 
     bg = "#aadaff")

# Second: add a Choropleth layer with the variable encoded in categories...
my_classes <- 8
my_method <-  "quantile"


choroLayer(
  x = my_geodata, 
  var = "X2019",
  method = my_method,
  nclass= my_classes,
  # col = carto.pal(pal1 = "sand.pal", n1 = nb_classes),
  col = heat.colors(my_classes),
  colNA	= "grey",
  border = "grey", 
  lwd = 0.5,
  legend.pos = c(40,-50), 
  legend.horiz = TRUE, 
  legend.title.txt = paste("Method:", my_method,",", my_classes,"classes"),
  add = TRUE
) 

# Third: add  layout
layoutLayer(title = paste(sdg_label, "in 2019"),
            col = "grey",
            frame = FALSE, 
            tabtitle = TRUE,
            posscale =  c(10, 40), 
            north = FALSE, # would be on top right, 
            )
# north arrow
north(pos = "topleft")


```


# Proportional symbol maps 
## Bubble charts with ggplot


```{r}
# we need to have the lat and logitude of each countries to be able to place the circles (points) 
world_data <- map_data("world") %>%
  select(region, lat, long) %>%
  group_by(region) %>%
  mutate(my_lat = median(lat, na.rm = TRUE), 
         my_long = median(long, na.rm = TRUE)) %>%
  distinct(region,my_lat, my_long, .keep_all = TRUE) %>%
  filter(
    long >50
  )
# Adding this information to my spatial data set
my_geodata_world <- merge(my_geodata, world_data, by.x= "NAME_LONG", by.y = "region")


my_geodata_world %>%
ggplot() + 
 geom_sf( color = "grey", fill = "white")+
 geom_point(aes(x=my_long, y=my_lat, size= X2019),
            color = SIAP.color ,
            stroke=F, 
            alpha=0.5) +

  ggtitle(sdg_label) + 
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_text(color = SIAP.color, size = 10))
```



### Caveat! {-} 

For *geom_point*, the size parameter scales **both** the x and y dimension of the point
proportionately. So a value that is double than another will have its surface multiplied by 4
Hence, the size of the points in R are proportional to its $radius^2$ . This is not a good way to practice!

```{r}
ggplot(data.frame(x     = 0, 
                  y     = 0, 
                  panel = c("A", "B", "C"), 
                  value = c(10, 20, 40))) + 
  geom_point(aes(x, y, size = value), color = SIAP.color) +
  coord_cartesian(xlim = c(-10, 10), ylim = c(-10,10)) +
  scale_y_continuous(breaks = c(-5, -2.5, 0, 2.5,5)) +
  scale_size_identity() +
  facet_wrap(.~panel) +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank())
```
 

### Solution: Compute the right value and plot the surface  {-}

```{r}
my_geodata_world %>%
  mutate(area_X2019 = sqrt(X2019))%>%
ggplot() + 
 geom_sf( color = "grey", fill = "white") +
 geom_point(aes(x=my_long, y=my_lat, size= area_X2019),
            color = SIAP.color ,
            alpha=0.5) +
  scale_size_area() +
  ggtitle(sdg_label) + 
  labs(fill = "Perc. of women \n(Indic. 5.5.1)")+
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_text(color = SIAP.color, size = 10))
```





## Proportional symbol maps with *cartography*


```{r}
# First: define the shape of the map
plot(st_geometry(my_geodata),  col= "#f0f0f0", border = "black",   bg = "#aadaff")

# Second: Plot symbols with choropleth coloration
my_symbol <- "circle"   # coud be "square" or "bar"
my_size <- 0.2

propSymbolsLayer(
  x = my_geodata, 
  var = "X2019",
  inches = my_size,
  symbols  = my_symbol, 
  border = "grey",
  col = "#E84923",
  lwd = 0.1,
  legend.pos = c(40,-50),
  legend.title.txt = paste0("Symbol: ", my_symbol,", size =", my_size)
) 

# Third: add  layout
layoutLayer(title = paste(sdg_label, "in 2019"),
            col = "grey",
            frame = FALSE, 
            tabtitle = TRUE,
            posscale =  c(10, 40), 
            north = FALSE, # would be on top right, 
            )
# north arrow
north(pos = "topleft")

```


```{r}
# First: define the shape of the map
plot(st_geometry(my_geodata),  col= "#f0f0f0", border = "black",   bg = "#aadaff")

# Second: Plot symbols with choropleth coloration
my_symbol <- "square"   # coud be "square" or "bar"
my_size <- 0.3

propSymbolsLayer(
  x = my_geodata, 
  var = "X2019",
  inches = my_size,
  symbols  = my_symbol, 
  border = "grey",
  col = SIAP.color,
  lwd = 0.1,
  legend.pos = c(40,-50),
  legend.title.txt = paste0("Symbol: ", my_symbol,", size =", my_size)
) 

# Third: add  layout
layoutLayer(title = paste(sdg_label, "in 2019"),
            col = "grey",
            frame = FALSE, 
            tabtitle = TRUE,
            posscale =  c(10, 40), 
            north = FALSE, # would be on top right, 
            )
# north arrow
north(pos = "topleft")

```


```{r}
# First: define the shape of the map
plot(st_geometry(my_geodata),  col= "#f0f0f0", border = "black",   bg = "#aadaff")

# Second: Plot symbols with choropleth coloration
my_symbol <- "bar"   # coud be "square" or "bar"
my_size <- 0.6

propSymbolsLayer(
  x = my_geodata, 
  var = "X2019",
  inches = my_size,
  symbols  = my_symbol, 
  border = "grey",
  col = "#E84923",
  lwd = 0.1,
  legend.pos = c(40,-50),
  legend.title.txt = paste0("Symbol: ", my_symbol,", size =", my_size)
) 

# Third: add  layout
layoutLayer(title = paste(sdg_label, "in 2019"),
            col = "grey",
            frame = FALSE, 
            tabtitle = TRUE,
            posscale =  c(10, 40), 
            north = FALSE, # would be on top right, 
            )
# north arrow
north(pos = "topleft")

```




### Mixing colors and size...

```{r}
library(lwgeom)

# First: define the shape of the map
plot(st_geometry(my_geodata),   col= "#f0f0f0",  border = "black",   bg = "#aadaff")

# Second: Plot symbols with choropleth coloration

my_classes <- 8
my_method <-  "quantile"

propSymbolsChoroLayer(
  x = my_geodata, 
  var = "X2019",
  var2 = "X2019",
  method = "equal",
  nclass= my_classes,
  inches = 0.2,
  col = heat.colors(my_classes),
  colNA	= "grey",
  border = "grey",
  lwd = 0.5,
  legend.var.pos = c(40,-50),
  legend.var.title.txt = paste("Method:", my_method,",", my_classes,"classes"), 
  legend.var2.pos =  "n"
) 

# Third: add  layout
layoutLayer(title = paste(sdg_label, "in 2019"),
            col = "grey",
            frame = FALSE, 
            tabtitle = TRUE,
            posscale =  c(10, 40), 
            north = FALSE, # would be on top right, 
            )
# north arrow
north(pos = "topleft")


```




```{r}

# First: define the shape of the map
plot(st_geometry(my_geodata),   col= "#f0f0f0",  border = "black",   bg = "#aadaff")

# Second: Plot symbols with choropleth coloration

my_classes <- 8
my_method <-  "quantile"

propSymbolsChoroLayer(
  x = my_geodata, 
  var = "X2019",
  var2 = "X2019",
  method = "equal",
  symbols = "square",
  nclass= my_classes,
  col = heat.colors(my_classes),
  colNA	= "grey",
  border = "grey",
  lwd = 0.5,
  legend.var.pos = c(40,-50),
  legend.var.title.txt = paste("Method:", my_method,",", my_classes,"classes"), 
  legend.var2.pos =  "n"
) 

# Third: add  layout
layoutLayer(title = paste(sdg_label, "in 2019"),
            col = "grey",
            frame = FALSE, 
            tabtitle = TRUE,
            posscale =  c(10, 40), 
            north = FALSE, # would be on top right, 
            )
# north arrow
north(pos = "topleft")


```






```{r}
knitr::knit_exit()
```






### hexbins

```{r, eval = FALSE}
library(cartography)
my_grid <- getGridLayer(
  x = my_geodata, 
  cellsize = 500 , 
  var = "X2019",
  type = "hexagonal"
)

plot(st_geometry(my_geodata),
     col= "#f0f0f0",
     border = "black", 
     bg = "#aadaff")

choroLayer(
  x = my_geodata, 
  var = "X2019",
  method = my_grid,
  nclass= my_classes,
  # col = carto.pal(pal1 = "sand.pal", n1 = nb_classes),
  col = heat.colors(my_classes),
  colNA	= "grey",
  border = "grey", 
  lwd = 0.5,
  legend.pos = c(40,-50), 
  legend.horiz = TRUE, 
  legend.title.txt = paste("Method:", my_method,",", my_classes,"classes"),
  add = TRUE
) 

# Third: add  layout
layoutLayer(title = paste(sdg_label, "in 2019"),
            col = "grey",
            frame = FALSE, 
            tabtitle = TRUE,
            posscale =  c(10, 40), 
            north = FALSE, # would be on top right, 
            )
# north arrow
north(pos = "topleft")



```




## Proportional symbols

```{r}
tmap_mode("plot")

tm_shape(my_geodata) +
   tm_borders("grey", lwd = .5) +
  # tm_text("ABBREV", size = "AREA") +
  tm_bubbles(size = "X2019", scale = 0.5) +
  tm_legend(show = TRUE)
```






```{r}
## regions Colors
#my_cols <- c("#00AFBB", "#E7B800", "#FC4E07") 
#my_cols <-c("#6b03fc", "#4287f5", "#69b3a2", "#03fca1", "#dffc03", "#E8E8E8") )
my_cols <- c(ggplotColours(n = 5), "#E8E8E8") 
my_cols <- c("#F8766D","#A3A500" , "#69b3a2","#00B0F6", "#E76BF3", "#E8E8E8")

library(RColorBrewer)
# Tree colors
my_highlow <- c('#fc8d59','#ffffbf','#91bfdb')
```



