---
title: "Pie Charts"
subtitle: "Why Pie Chart is a nightmare"
author: "Christophe Bontemps (UN SIAP)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: lumen
  html_notebook: default
  pdf_document: 
    keep_tex: yes  ## keep images in folder + latex
---


```{r setup, include=FALSE}
# Remeber to disable png for Word export...
knitr::opts_chunk$set( message = FALSE, warning = FALSE, 
                       results =FALSE, echo = FALSE,
                       dev="png", 
                       dev.args=list(type="cairo"), dpi=96)

# My colors:
SIAP.color <- "#0385a8"

# Function used to recreate ggplot standard colors
ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

```

```{r libraries}

library(tidyverse)
library(viridis)
library(ggthemes)


library(hrbrthemes)
library(viridis)
library(patchwork)

```

## Data set {-}
Let's say we have observed 5 categories  from 3 countries
```{r Data}
# create 3 data frame:
data1 <- data.frame( name=letters[1:5], value=c(17,18,20,22,24) )
data2 <- data.frame( name=letters[1:5], value=c(20,18,21,20,20) )
data3 <- data.frame( name=letters[1:5], value=c(24,23,21,19,18) )
```

## Classical pie chart by country {-}  
```{r SinglePie}
# From-data-to viz https://www.data-to-viz.com/caveat/pie.html
myPalette <-  c('#fdbf6f', '#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c')
myPalette1 <-  c('#fdbf6f', '#a6cee3','#b2df8a',"grey",'#e31a1c')
myPalette2 <-  c('#fdbf6f', '#a6cee3','pink',"grey",SIAP.color)
myPalette3 <- c(SIAP.color, "grey", "orange", "darkred", "black")
#myPalette4 <- c("lightyellow", "yellow",'#b2df8a','#33a02c','#fb9a99','#e31a1c', "orange")



plot_pie <- function(data, vec){

ggplot(data, aes(x="name", y=value, fill=name)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0, direction = -1) +
  scale_fill_manual("Categories",values=myPalette2) +
  geom_text(aes(y = vec, label = rev(name), size=4, color=c( "white", rep("black", 4)))) +
  scale_color_manual(values=c("black", "white")) +
  theme_minimal() +
  theme(
    legend.position="none",
    plot.title = element_text(size=14),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    legend.margin=unit(0, "null")
  ) +
  xlab("") +
  ylab("")
  
}

# First country
plot_pie(data1, c(10,35,55,75,93))

```

# All countries compared with Pies {-}


```{r 3Pies}
p1 <- plot_pie(data1, c(10,35,55,75,93))
p2 <- plot_pie(data2, c(10,35,53,75,93))
p3  <- plot_pie(data3, c(10,29,50,75,93))
p1 + p2 + p3 + 
 plot_annotation(tag_levels = 'I')
```

# All countries compared with Bars {-}


```{r 3barsV}
# A function to make vertical barplots
plot_bar_V <- function(data){
  ggplot(data, aes(x=name, y=value, fill=name)) +
    geom_bar( stat = "identity") +
    scale_fill_manual("Categories",values=myPalette2) +
    scale_color_manual(values=c("black", "white")) +
    theme_minimal() +
    theme(
      legend.position="none",
      plot.title = element_text(size=14),
      panel.grid = element_blank(),
    ) +
    ylim(0,25) +
    xlab("") +
    ylab("")
}

# Make 3 barplots
bar1v <- plot_bar_V(data1)
bar2v <- plot_bar_V(data2)
bar3v <- plot_bar_V(data3)

# Put them together with patchwork
bar1v + bar2v + bar3v + 
 plot_annotation(tag_levels = 'I')
```


# All countries compared with *horizontal* bars {-}


```{r 3bars}
# A function to make vertical barplots
plot_bar_H <- function(data){
  ggplot(data, aes(x=fct_rev(name), y=value, fill= name)) +
    geom_bar( stat = "identity") +
    scale_fill_manual("Categories",values=myPalette2) +
    scale_color_manual(values=c("black", "white")) +
    coord_flip()+
    theme_minimal() +
    theme(
      legend.position="none",
      plot.title = element_text(size=14),
      panel.grid = element_blank(),
    ) +
    ylim(0,25) +
    xlab("") +
    ylab("")
}

# Make 3 barplots
bar1h <- plot_bar_H(data1)
bar2h <- plot_bar_H(data2)
bar3h <- plot_bar_H(data3)

# Put them together with patchwork
bar1h  /bar2h  / bar3h + 
 plot_annotation(tag_levels = 'I')
```

#  Comparisons {-}

```{r Compare1}
(p1+p1+p3)  / (( bar1v + bar2v+ bar3v)  +  plot_layout(tag_level = 'new')) +
  plot_annotation(tag_levels = list(c('I', 'II', "III"), c('', '', "III")))


```


```{r Compare2}
(p1 + bar1h) /
(p2 + bar2h) /
(p3 + bar3h)

```

# Multiple categories pie charts




```{r}
library(tidyverse)

# Defining my palettes
myPalette11 <- c(
  '#b2df8a',  # light green
  '#cab2d6',  # lavender
  '#ffff99',  # pastel yellow
  '#fdbf6f',  # light orange (repeated if intentional, can adjust)
  '#fb9a99',  # salmon pink
  '#1f78b4',  # steel blue
  'grey',  # lighter red
  '#b2df8a',  # pastel green
  '#fdbf6f',  # light orange
  '#a6cee3',  # soft blue
  'pink'     # soft pink
)

myPalette4 <-  c(
'pink',  '#a6cee3','#fdbf6f',"lightgrey"
)
               
myPalette4I <-  c(
"lightgrey",  '#fdbf6f', '#a6cee3','pink'
)
               
myPalette4Ordered <-  c(
'pink', "lightgrey",  '#a6cee3', '#fdbf6f'
)
               

```


```{r}
# Data generation 
set.seed(123)
product_data <- tibble(
  category = c("Rice",  "Textiles", 
               "Tea", "Rubber", "Spices", "Coconuts", 
               "Fish", "Handicrafts", "Fruits", "Bamboo", "Electronics"),
  value = c(35, 25, 20, runif(8, 1, 5))
) %>%
  arrange(desc(value)) %>%
  mutate(
    percent = value / sum(value),
    label = paste0(category, "\n", round(100 * percent), "%"),
    category = fct_reorder(category, value)
  )

```


### Ploting all categories


```{r}
product_data %>%
  ggplot(aes(x = "", y = value, fill = category)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  scale_fill_manual("Categories",values=myPalette11) +
  geom_text(aes(label = category),
            position = position_stack(vjust = 0.5),
            size = 4,
            color = "black", alpha = 0.8) +
  theme_void() +
#  ggtitle("Pie Chart with 11  Product Categories (Ordered)") +
  theme(legend.position = "none")

```



### Ploting with aggreagted categories


```{r}
# Choose top 3 as "main"
main_categories <- product_data %>%
  top_n(3, value) %>%
  pull(category)

# Recode other categories as "Misc"
collapsed_data <- product_data %>%
  mutate(category = if_else(category %in% main_categories, category, "Misc")) %>%
  group_by(category) %>%
    summarise(value = sum(value), .groups = "drop") %>%
  ungroup() %>%
   mutate(
  category = fct_relevel(category, "Misc", "Tea", "Textiles", "Rice")
  )
   

# Pie chart with collapsed categories
collapsed_data %>%
  ggplot(aes(x = "", y = value, fill = category)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = myPalette4I) +
  geom_text(aes(label = category),
            position = position_stack(vjust = 0.5),
            size = 5,
            color = "black") +
  theme_void() +
 # ggtitle('Pie Chart with only 4 categories, including a "Misc" category ') +
  theme(legend.position = "none")
```
## Alternative to pie chart

```{r}
collapsed_data %>%
  ggplot(aes(x= "" , y = value, fill = category)) +
  geom_bar(position="stack", stat="identity", width = 0.2) +
  scale_fill_manual(values = myPalette4I) +
  geom_text(aes(label = category),
            position = position_stack(vjust = 0.5),
            size = 5,
            color = "black") +
  theme_void() +
#  ggtitle('Stacked bar as alternative to Pie Chart with only 4 categories') +
  theme(legend.position = "none")
```



```{r}

collapsed_data %>%
   mutate(category = fct_reorder(category, value, .desc = TRUE)) %>%
   ggplot(aes(x = factor(1), y = value, fill = category)) +
  geom_bar(stat = "identity", 
           width = 0.8, 
           position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = myPalette4Ordered) +
  geom_text(aes(label = category),
            position = position_dodge(width = 0.8),
            vjust = +3.5,
            size = 5,
            color = "black") +
  theme_void() +
  theme(legend.position = "none") +
  coord_fixed(ratio = 0.01) +
  theme_void() +
#  ggtitle('Stacked bar as alternative to Pie Chart with only 4 categories') +
  theme(legend.position = "none")
```



# Comparing areas (circles)

```{r}
library(tidyverse)

# Data for one circle
circle1 <- tibble(
  angle = seq(0, 2 * pi, length.out = 200),
  x = 9 * cos(angle),
  y = 9 * sin(angle)
)

# Compute area
area1 <- pi * 9^2

# Plot with annotation
ggplot(circle1, aes(x, y)) +
  geom_polygon(fill = "grey70", color = "grey50") +
  annotate("text", x = 0, y = 0, label = paste0("Area ≈ ", round(area1, 1)), size = 6) +
  coord_fixed() +
  theme_void() +
  ggtitle("One Filled Disk (radius = 9)")



```
## 2 discks

```{r}
# Data for two disks
circle2 <- tibble(
  angle = seq(0, 2 * pi, length.out = 200)
) %>%
  expand_grid(center_x = c(-7, 7)) %>%
  mutate(
    x = center_x + 6 * cos(angle),
    y = 6 * sin(angle),
    group = as.factor(center_x)
  )

# Compute area for each
area2 <- pi * 6^2

# Plot with annotations
ggplot(circle2, aes(x, y, group = group)) +
  geom_polygon(fill = "grey70", color = "grey50") +
  annotate("text", x = -7, y = 0, label = paste0("Area ≈ ", round(area2, 1)), size = 5) +
  annotate("text", x = 7, y = 0, label = paste0("Area ≈ ", round(area2, 1)), size = 5) +
  coord_fixed() +
  theme_void() +
  ggtitle("Two Filled Disks (radius = 6)")


```



