---
title: "Creating a Statistical Map"
subtitle: "A step-by-step case study with Bangladesh data "
author: "Christophe Bontemps (UN SIAP)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: 
    keep_tex: yes  ## keep images in folder + latex
  html_notebook: default
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: lumen
---
```{r setup, include=FALSE}
# Remeber to disable png for Word export...
knitr::opts_chunk$set( message = FALSE, warning = FALSE, 
                       results =FALSE, echo = FALSE,
                       dev="png", 
                       dev.args=list(type="cairo"), dpi=96)

# My colors:
SIAP.color <- "#0385a8"

```

```{r CartoPackage}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_gpkg <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_gpkg, quiet = TRUE)
# Plot the municipalities
plot(st_geometry(mtq), col="darkseagreen3", border="darkseagreen4",  
     bg = "lightblue1", lwd = 0.5)
# Plot symbols with choropleth coloration
propSymbolsLayer(
  x = mtq, 
  var = "POP", 
  inches = 0.25, 
  col = "brown4",
  legend.pos = "topright",  
  legend.title.txt = "Total population"
)
# layout
layoutLayer(title="Population & Wealth in Martinique, 2015", 
            author = "cartography 2.1.3", 
            sources = "Sources: Insee and IGN, 2018", 
            scale = 5, tabtitle = TRUE, frame = FALSE)
# north arrow
north(pos = "topleft")
```







* * *
# 1. Introduction to integrating household survey and geospatial data

We need a minimum of organisation in the data and code folders, as well as some R packages.

```{r packages}
# GIS packages
library(raster) ## for reading "RASTER" files
library(rgdal)  ## for reading "shapefiles"
library(sp)     ## for adjusting CRS in 


# Tidy data management packages
library(dplyr)
library(data.table)


# Plotting packages
library(ggplot2)
library(RColorBrewer)

# Nice presentation of results
library(knitr)
library(papeR)
```



```{r}
# --- Change to YOUR project folder HERE  ---- #

source_folder<-"c:/Chris/UN-ESCAP/SIAP E-learning/TechnicalStuff/UNWomen-GISBangladesh/"


# Specific sub-folder for data storage

shapedata_folder<-paste(source_folder, "dhsdata/BDGE71FL", sep="")  
geodata_folder<-paste(source_folder, "geodata/", sep="")
data_folder<-paste(source_folder,"Data/",sep="")

# this is where all saved .Rda go and can be loaded when needed
output_folder<-paste(source_folder, "CreatedData/" , sep="")  
```


# 2. Understanding child marriage using geo-covariates

## The DHS survey


```{r}
# Reading DHS survey data 
merged1<-read.csv(file = 'Data/bangladesh.csv')  # reading DHS Bangladesh 2014
merged1$Age<-as.numeric(merged1$Age)

#Computing the proportion of getting married before 15 by cluster
cluster_average<-aggregate(Before15~DHSCLUST,
                           data=merged1,
                           FUN=mean)  
```

# 3. Integrating DHS and geospatial data

## Reading the DHS Shapefile 

```{r}
# Reading DHS Shapefile 
dhsShapeData<-readOGR(shapedata_folder, "BDGE71FL") # Reads the shapefile in DHS
shapedata<-dhsShapeData@data                        # Reads the data part 
shapefile_df <- fortify(shapedata)
shapedata<-shapedata[shapedata$LATNUM>0, ]          # Drops negative Latnum 
```

### Cluster locations by urban and rural from the shapefile (DHS Bangladesh 2014) {-}

We can use the *latitude* and *longitude* of each observation  to draw a "location map" of the clusters. This generates the **Figure 3** of the document.  

>This is not really "a map", but only points with latitude and longitude defined represented on a grid


```{r shapefile}
# Now the shapefile can be plotted, as points 
# In the aesthetics part of the ggplot we need long, lat,
#                       and we use group for Urban or Rural.
map <-ggplot(shapedata,  aes(x = LONGNUM, y = LATNUM, color = factor(URBAN_RURAL))) +
  geom_point(color = SIAP.color, alpha = 0.5) +
 ggtitle("Shapefile Cluster Location on and X-Y grid") + 
         labs(x=  "X-axis used as Longitude" ,
              y = "Y-axis used as Latitude")

# Using the ggplot2 function coord_map will make things look better
# and it will also let you change the projection.
map_projected <- map +
  coord_map()+
  theme_minimal()

map_projected

```



```{r}
# Reading geographic data - access to cities
accessData<-raster(paste(geodata_folder,
                         "accessibility_to_cities_2015.tif",
                         sep=""))
accessData
#  We need to provide the same  projection reference 
#  We use sp::spTransform to tell wich CRS is used.

dhsShapeData2 <- spTransform(dhsShapeData, accessData@crs)
```
```{r shapefileBinary}
# Now the shapefile can be plotted, as points 
# In the aesthetics part of the ggplot we need long, lat,
#                       and we use group for Urban or Rural.
map <- ggplot() +
  geom_point(data = shapedata,
              aes(x = LONGNUM, y = LATNUM, color = factor(URBAN_RURA)),
              size = .6) +
 ggtitle("Shapefile Cluster Location on and X-Y grid") + 
         labs(x=  "X-axis used as Longitude" ,
              y = "Y-axis used as Latitude")

# Using the ggplot2 function coord_map will make things look better
# and it will also let you change the projection.
map_projected <- map +
  coord_map()+
  theme_minimal()

map_projected
```

## Reading raster files 
 
```{r}
# Reading geographic data - access to cities
accessData<-raster(paste(geodata_folder,
                         "accessibility_to_cities_2015.tif",
                         sep=""))
accessData
```


```{r MapLocation}
# we need to restrict the raster to Bangladesh at some point... 

accessData.bd<- crop(accessData, extent(88.01057, 92.67723, 20.74111, 26.63278))

plot(accessData.bd, breaks= c(-10, 0, 50, 100, 200, 500 ),
     col = c("white", "#ace8b6", "#9fcaf5", "#cf9ef7","#f79ed2", "#f79ec5", "#f02e5e") ,
    # main="Location of observations",
     axes=FALSE, box=FALSE,
     legend=FALSE)

#I add the cluster points (overlap on Bangladesh map... ) 
set.seed(2512)
MySample <- sample_n(shapedata, 100)
points(x=MySample$LONGNUM, y=MySample$LATNUM, 
        type="p", 
        cex=0.8, pch=21, bg=1, 
        col="red")

```

## Playing

```{r MapSymbols}
# we need to restrict the raster to Bangladesh at some point... 

accessData.bd<- crop(accessData, extent(88.01057, 92.67723, 20.74111, 26.63278))

plot(accessData.bd, breaks= c(-10, 0, 50, 100, 200, 500 ),
     col = c("white", "#ace8b6", "#9fcaf5", "#cf9ef7","#f79ed2", "#f79ec5", "#f02e5e") ,
     #main = "Statistical map of an indicator",
     #sub = "Visual encoding with symbols",
     axes=FALSE, box=FALSE,
     legend=FALSE)

#I add the cluster points (overlap on Bangladesh map... ) 


points(x=MySample$LONGNUM, y=MySample$LATNUM,
        type= "p" ,
        cex= c(1.2), 
        pch= c(4, 15, 16, 17, 20), 
        bg=1, 
        col="purple", 
        legand =TRUE)

```




```{r MapSize }
plot(accessData.bd, breaks= c(-10, 0, 50, 100, 200, 500 ),
     col = c("white", "#ace8b6", "#9fcaf5", "#cf9ef7","#f79ed2", "#f79ec5", "#f02e5e") ,
    # main = "Statistical map of an indicator",
    # sub = "Visual encoding with size",
     axes=FALSE, box=FALSE,
     legend=FALSE)

#I add the cluster points (overlap on Bangladesh map... ) 
 points(x=MySample$LONGNUM, y=MySample$LATNUM,
        type= "p" ,
        cex= c(0.4,0.8, 1, 1.8), 
        pch= 16, 
        bg=1, 
        col="purple", 
        legand =TRUE)

```



```{r MapSizeColors}
plot(accessData.bd, breaks= c(-10, 0, 50, 100, 200, 500 ),
     col = c("white", "#ace8b6", "#9fcaf5", "#cf9ef7","#f79ed2", "#f79ec5", "#f02e5e") ,
    # main = "Statistical map of an indicator",
    # sub = "Visual encoding with size and color",
     axes=FALSE, box=FALSE,
     legend=FALSE)

#I add the cluster points (overlap on Bangladesh map... ) 
 points(x=MySample$LONGNUM, y=MySample$LATNUM,
        type= "p" ,
        cex= c(0.4,0.8, 1.2, 1.8), 
        pch= 16, 
        bg=1, 
        col= c( "green", "red", "purple", "darkgray"), 
        legand =TRUE)

```

